---
title: "API Reference"
---


```{r, echo=FALSE}
library(sicher)
```


## Core Functions

### `create_type(name, checker)`

Creates a new type with a name and validation function.

**Parameters:**
- `name`: Character string for error messages
- `checker`: Function that returns TRUE/FALSE for validation

**Example:**
```{r}
Positive <- create_type("positive", function(x) is.numeric(x) && all(x > 0))
```

### `create_list_type(type_spec)`

Creates a type for lists with specific field requirements.

**Parameters:**
- `type_spec`: Named list where names are field names and values are types

**Example:**
```{r}
Person <- create_list_type(list(name = String, age = Numeric))
```

### `create_dataframe_type(col_spec)`

Creates a type for data frames with column specifications.

**Parameters:**
- `col_spec`: Named list where names are column names and values are types

**Example:**
```{r}
Table <- create_dataframe_type(list(id = Integer, name = String))
```

### `ListOf(element_type)`

Creates a type for homogeneous lists.

**Parameters:**
- `element_type`: Type that each list element must satisfy

**Example:**
```{r}
Numbers <- ListOf(Numeric)
```

## Type Modifiers

### `Scalar(type)`

Requires exactly one element (length 1).

**Example:**
```{r}
SingleValue <- Scalar(Numeric)
```

### `Readonly(type)`

Prevents reassignment after initial value.

**Example:**
```{r}
Constant <- Readonly(String)
```

### `Optional(type)`

Allows NULL in addition to the base type.

**Example:**
```{r}
MaybeString <- Optional(String)
```

## Operators

### `%:%`

Type annotation operator.

**Usage:**
```r
variable %:% Type
```

### `%<-%`

Typed assignment operator.

**Usage:**
```r
variable %:% Type %<-% value
```

### `|`

Union operator for combining types.

**Example:**
```r
StringOrNumber <- String | Numeric
```

### `[size]`

Size constraint operator.

**Example:**
```r
ThreeNumbers <- Numeric[3]
```

## Built-in Types

### Primitive Types

- `Integer`: Integer vectors
- `Double`: Double-precision numeric vectors
- `Numeric`: Any numeric vectors
- `String`: Character vectors
- `Bool`: Logical vectors
- `List`: List objects
- `Any`: Accepts any value
- `Null`: Only NULL values
- `Function`: Function objects
- `DataFrame`: Data frame objects

## Error Handling

Type errors provide detailed context:

```{r, error=TRUE}
x %:% Numeric %<-% "text"
```

Error messages include:
- Expected type
- Actual type
- Variable name (when available)
- Value preview (for debugging)
