---
title: "API Reference"
---

```{r, echo=FALSE}
library(sicher)
```

## Core Functions

### `typed_function(fn, params, .return)`

Wraps a function with runtime type checks for its parameters and, optionally, its return value.

**Parameters:** - `fn`: The function to wrap. Its formals are preserved so callers use the same signature. - `params`: A named list mapping parameter names to their types (e.g. `list(x = Numeric, y = String)`). Only listed parameters are checked. Defaults to an empty list (no parameter checking). - `.return`: Optional return type. When `NULL` (the default), the return value is not checked. Accepts any `sicher_type` or `sicher_union`.

**Returns:** A function with S3 class `"sicher_typed_function"` that validates types on every call and delegates argument passing to `fn` unchanged.

**Example:**

```{r}
add <- typed_function(
  function(x, y) x + y,
  params  = list(x = Numeric, y = Numeric),
  .return = Numeric
)

greet <- typed_function(
  function(name, title = NULL) {
    if (is.null(title)) paste("Hello,", name)
    else paste("Hello,", title, name)
  },
  params = list(name = String, title = Optional(String))
)
```

### `create_type(name, checker)`

Creates a new type with a name and validation function.

**Parameters:** - `name`: Character string for error messages - `checker`: Function that returns TRUE/FALSE for validation

**Example:**

```{r}
Positive <- create_type("positive", function(x) is.numeric(x) && all(x > 0))
```

### `create_list_type(type_spec)`

Creates a type for lists with specific field requirements.

**Parameters:** - `type_spec`: Named list where names are field names and values are types

**Example:**

```{r}
Person <- create_list_type(list(name = String, age = Numeric))
```

### `create_dataframe_type(col_spec)`

Creates a type for data frames with column specifications.

**Parameters:** - `col_spec`: Named list where names are column names and values are types

**Example:**

```{r}
Table <- create_dataframe_type(list(id = Integer, name = String))
```

### `Enum(...)`

Creates a type that only accepts values from a fixed set.

**Parameters:** - `...`: Allowed scalar values, or a single atomic vector containing the allowed values

**Example:**

```{r, error=TRUE}
status %:% Enum("draft", "published", "archived") %<-% "draft"
priority %:% Enum(1, 2, 3) %<-% c(1, 2)

status <- "published"
status <- "Ho!"

priority <- c(3, 2, 1)
priority <- c(3, 2, 1, 100000)

```

`Enum(...)` builds a regular `sicher_type`, so it works anywhere another type works:

```{r, error=TRUE}
status %:% Enum("draft", "published", "archived") %<-% "draft"
review <- typed_function(
  function(state) paste("State:", state),
  params = list(state = Enum("draft", "published", "archived")),
  .return = String
)

review("published")
review("hola")
```

### `Literal(...)`

Creates a type that only accepts exact scalar atomic literals, inspired by TypeScript literal types.

**Parameters:** - `...`: Allowed scalar atomic literal values

**Example:**

```{r, error=TRUE}
direction %:% Literal("left", "right") %<-% "left"
direction <- "right"
direction <- c("left", "right")

status_code %:% Literal(200, 404) %<-% 200
status_code <- 404
status_code <- 200L
```

`Literal(...)` is stricter than `Enum(...)`: it accepts only length-1 atomic values and compares them exactly, so `200` and `200L` are different literals.

```{r, error=TRUE}
move <- typed_function(
  function(direction) paste("Moving", direction),
  params = list(direction = Literal("left", "right")),
  .return = String
)

move("left")
move("up")
```

### `ListOf(element_type)`

Creates a type for homogeneous lists.

**Parameters:** - `element_type`: Type that each list element must satisfy

**Example:**

```{r}
Numbers <- ListOf(Numeric)
```

## Type Modifiers

### `Scalar(type)`

Requires exactly one element (length 1).

**Example:**

```{r}
SingleValue <- Scalar(Numeric)
```

### `Readonly(type)`

Prevents reassignment after initial value.

**Example:**

```{r}
Constant <- Readonly(String)
```

### `Optional(type)`

Allows NULL in addition to the base type.

**Example:**

```{r}
MaybeString <- Optional(String)
```

## Operators

### `%:%`

Type annotation operator.

**Usage:**

``` r
variable %:% Type
```

### `%<-%`

Typed assignment operator.

**Usage:**

``` r
variable %:% Type %<-% value
```

### `|`

Union operator for combining types.

**Example:**

``` r
StringOrNumber <- (String | Numeric)
```

### `[size]`

Size constraint operator.

**Example:**

``` r
ThreeNumbers <- Numeric[3]
```

## Built-in Types

### Primitive Types

-   `Integer`: Integer vectors
-   `Double`: Double-precision numeric vectors
-   `Numeric`: Any numeric vectors
-   `String`: Character vectors
-   `Bool`: Logical vectors
-   `List`: List objects
-   `Any`: Accepts any value
-   `Null`: Only NULL values
-   `Function`: Function objects
-   `DataFrame`: Data frame objects

## Error Handling

Type errors provide detailed context:

```{r, error=TRUE}
x %:% Numeric %<-% "text"
```

Error messages include: - Expected type - Actual type - Variable name (when available) - Value preview (for debugging)


## Global Options

The behavior of **sicher** can be controlled globally using R’s `options()` mechanism. These options allow you to switch between strict type enforcement and a fully disabled mode depending on your use case.

```{r, error=TRUE}
my_var_x %:% Numeric %<-% 10   # validated and bound with type enforcement

my_var_x <- "string" # trigger an error
```


```{r}
options(sicher.mode = "off")
my_var_y %:% Numeric %<-% 10 
my_var_y <- "string" # nothing happens
```

It can also be temporarily scoped:
```{r}
withr::with_options(
  list(sicher.mode = "off"),
  {
    my_var_y %:% Numeric %<-% 10
    my_var_y <- "string"
  }
)
```

```{r, echo=FALSE}
options(sicher.mode = "on")
```

