---
title: "Features"
---

## Core Features

### Built-in Types

sicher provides a comprehensive set of built-in types:

```{r}
library(sicher)

# Basic types
x %:% Integer %<-% 42L
y %:% Double %<-% 3.14
z %:% Numeric %<-% 100
name %:% String %<-% "Alice"
flag %:% Bool %<-% TRUE
items %:% List %<-% list(1, 2, 3)
anything %:% Any %<-% "could be anything"
nothing %:% Null %<-% NULL
func %:% Function %<-% function(x) x + 1
df %:% DataFrame %<-% data.frame(a = 1:3, b = letters[1:3])
```

### Type Modifiers

#### Scalar Types

Enforce single values (vectors of length 1):

```{r}
#| error: true
# This works
single_value %:% Scalar(Numeric) %<-% 42

# This fails
single_value <- c(1, 2, 3)
```

#### Readonly Types

Prevent reassignment after initial value:

```{r}
#| error: true

# Set a readonly constant
PI %:% Readonly(Double) %<-% 3.14159

PI <- 3.0 # This fails
```

#### Optional Types

Allow NULL values in addition to the base type:

```{r}
#| error: true

PI <- 3.0

# Optional string (can be string or NULL)
middle_name %:% Optional(String) %<-% NULL
middle_name <- "Marie"  # Also OK

middle_name <- 123 # This fails
```

### Vector Size Types

Specify exact vector lengths:

```{r}
#| error: true

# Exactly 3 numeric values
coords %:% Numeric[3] %<-% c(1, 2, 3)

coords <- c(4, 5, 6) # This works

coords <- c(1, 2)  # Too short
```

### Union Types

Allow multiple types using the `|` operator:

```{r}
#| error: true

# Accept either string or numeric
id %:% (String | Numeric) %<-% "user123"
id <- 456  # Also OK

id <- TRUE
```

### Structured List Types

Define object-like structures for lists:

```{r}
#| error: true

# Define a Person type
Person <- create_list_type(list(
  name = String,
  age = Numeric,
  email = Optional(String)
))

# Use it
person %:% Person %<-% list(
  name = "Alice",
  age = 30,
  email = "alice@example.com"
)

# Missing required field fails
person <- list(name = "Bob")

# Extra fields not allowed fail
person <- list(
    name = "Charlie",
    age = 25,
    email = "charlie@example.com",
    extra = "not allowed"
)
```

### Data Frame Types

Define schemas for data frames:

```{r}
#| error: true

# Define a user table schema
UserTable <- create_dataframe_type(list(
  id = Integer,
  username = String,
  active = Bool,
  last_login = Optional(String)
))

# Valid data frame
users %:% UserTable %<-% data.frame(
  id = 1:3,
  username = c("alice", "bob", "charlie"),
  active = c(TRUE, FALSE, TRUE)
)

# Missing required column fails
users <- data.frame(
  username = c("alice", "bob"),
  active = c(TRUE, FALSE)
)

users <- data.frame(
  id = 1:2,
  username = c("alice", "bob"),
  active = c("yes", "no")  # Should be logical
)
```

### Homogeneous List Types

For lists of similar items (like API responses):

```{r}
#| error: true

# Define item structure
TodoItem <- create_list_type(list(
  userId = Numeric,
  id = Numeric,
  title = String,
  completed = Bool
))

# List of todo items
TodoList <- ListOf(TodoItem)

# Valid list
todos %:% TodoList %<-% list(
  list(userId = 1, id = 1, title = "Task 1", completed = FALSE),
  list(userId = 1, id = 2, title = "Task 2", completed = TRUE)
)

todos <- list(
  list(userId = 1, id = 1, title = "Task 1", completed = FALSE),
  list(wrong = "structure")
)
```

### Combining Features

All features can be combined:

```{r}
#| error: true

# Complex nested structure
Company <- create_list_type(list(
  name = String,
  employees = ListOf(create_list_type(list(
    name = String,
    salary = Numeric,
    department = Optional(String)
  ))),
  founded = Numeric[3]  # [year, month, day]
))

company %:% Company %<-% list(
  name = "Tech Corp",
  employees = list(
    list(name = "Alice", salary = 75000, department = "Engineering"),
    list(name = "Bob", salary = 65000)  # department optional
  ),
  founded = c(2020, 1, 15)
)
```

## Advanced Usage

### Custom Types

Create your own types using `create_type()`:

```{r}
#| error: true

# Positive number type
Positive <- create_type("positive", function(x) {
  is.numeric(x) && all(x > 0)
})

value %:% Positive %<-% 5

value <- -1
```

### Type Composition

Build complex types from simpler ones:

```{r}
#| error: true

# Email type
Email <- create_type("email", function(x) {
  is.character(x) && length(x) == 1 &&
  grepl("^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$", x)
})

# Person with email
PersonWithEmail <- create_list_type(list(
  name = String,
  email = Email,
  age = Numeric
))

person %:% PersonWithEmail %<-% list(
  name = "Alice",
  email = "alice@example.com",
  age = 30
)

person %:% PersonWithEmail %<-% list(
    name = "Bob",
    email = "invalid-email",
    age = 25
)
```

## Error Messages

`sicher` provides clear, contextual error messages:

```{r, error=TRUE}
#| error: true

# Type mismatch
x %:% Numeric %<-% "not a number"
```

```{r, error=TRUE}
# Missing required field
PersonType <- create_list_type(list(name = String, age = Numeric))
p %:% PersonType %<-% list(name = "Alice")
```

```{r, error=TRUE}
# Wrong vector length
vec %:% Numeric[3] %<-% c(1, 2)
```


