---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sicher 🔐

<!-- badges: start -->
<!-- badges: end -->

**sicher** (German for *safe* or *certain*) is an R package that brings runtime type safety to R programming — inspired by TypeScript for JavaScript. Declare types for your variables and have them enforced automatically on every assignment, catching type errors early and making your code more robust and self-documenting. 🛡️

## 📦 Installation

Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("feddelegrand7/sicher")
```

## 🚀 Quick Start

```{r, error = TRUE}
library(sicher)

# Annotate a variable with a type, then assign a value
name %:% String  %<-% "Alice"
age  %:% Numeric %<-% 30

# The type is enforced on every subsequent assignment
age <- "thirty"   # Error: Type error
```

## ✨ Core Features

### 🧩 Built-in Types

sicher ships with a complete set of primitive and container types:

| Type        | Checks                          |
|-------------|---------------------------------|
| `Integer`   | `is.integer()`                  |
| `Double`    | `is.double()`                   |
| `Numeric`   | `is.numeric()`                  |
| `String`    | `is.character()`                |
| `Bool`      | `is.logical()`                  |
| `List`      | `is.list()`                     |
| `DataFrame` | `is.data.frame()`               |
| `Function`  | `is.function()`                 |
| `Any`       | always passes                   |
| `Null`      | `is.null()`                     |

```{r}
x    %:% Integer   %<-% 42L
y    %:% Double    %<-% 3.14
flag %:% Bool      %<-% TRUE
df   %:% DataFrame %<-% data.frame(a = 1:3)
```

### ⚙️ Operators

| Operator | Purpose |
|----------|---------|
| `%:%`    | Annotate a variable with a type |
| `%<-%`   | Assign a value (type-checked)   |

### 🔧 Type Modifiers

#### 1️⃣ `Scalar()` — single-element values

```{r, error = TRUE}
single %:% Scalar(Numeric) %<-% 42
single <- c(1, 2, 3)   # Error: length > 1
```

#### 🔒 `Readonly()` — immutable variables

```{r, error = TRUE}
PI %:% Readonly(Double) %<-% 3.14159
PI <- 3.0   # Error: cannot reassign readonly variable
```

#### ❓ `Optional()` — nullable values

```{r, error = TRUE}
middle_name %:% Optional(String) %<-% NULL   # OK
middle_name <- "Marie"                        # Also OK
middle_name <- 123                            # Error: not string or null
```

### 🔀 Union Types

Accept more than one type with `|`:

```{r, error = TRUE}
id %:% (String | Numeric) %<-% "user123"
id <- 456   # Also OK
id <- TRUE  # Error: not string or numeric
```

### 📏 Size-constrained Vectors

Append `[n]` to any type to require an exact vector length:

```{r, error = TRUE}
coords %:% Numeric[3] %<-% c(1, 2, 3)
coords <- c(4, 5, 6)   # OK — same length
coords <- c(1, 2)      # Error: wrong length
```

### 📋 Structured List Types

Define object-like schemas with `create_list_type()`:

```{r, error = TRUE}
Person <- create_list_type(list(
  name  = String,
  age   = Numeric,
  email = Optional(String)   # nullable field
))

person %:% Person %<-% list(name = "Alice", age = 30, email = "alice@example.com")

person <- list(name = "Bob")   # Error: missing required field 'age'
```

### 🗄️ Data Frame Schemas

Validate column names and types with `create_dataframe_type()`:

```{r, error = TRUE}
UserTable <- create_dataframe_type(list(
  id       = Integer,
  username = String,
  active   = Bool
))

users %:% UserTable %<-% data.frame(
  id       = 1:2,
  username = c("alice", "bob"),
  active   = c(TRUE, FALSE)
)

# Wrong column type fails immediately
users <- data.frame(
  id       = c("1", "2"),   # Error: id must be integer
  username = c("alice", "bob"),
  active   = c(TRUE, FALSE)
)
```

### 📦 Homogeneous Lists with `ListOf()`

Validate every element of a list against the same type:

```{r, error = TRUE}
TodoItem <- create_list_type(list(
  id        = Numeric,
  title     = String,
  completed = Bool
))

TodoList <- ListOf(TodoItem)

todos %:% TodoList %<-% list(
  list(id = 1, title = "Buy milk",  completed = FALSE),
  list(id = 2, title = "Read book", completed = TRUE)
)

todos <- list(
  list(id = 1, title = "Buy milk", completed = FALSE),
  list(wrong = "shape")   # Error: element does not match TodoItem
)
```

### 🛠️ Custom Types

Use `create_type()` to define your own validator with any predicate function:

```{r, error = TRUE}
Positive <- create_type("positive", function(x) is.numeric(x) && all(x > 0))

value %:% Positive %<-% 5
value <- -1   # Error
```

## 🔤 Typed Functions

Use `typed_function()` to wrap any function with runtime type checks on its
parameters and, optionally, its return value — a typed function signature for R:

```{r, error = TRUE}
# Basic typed function — checks params and return type
add <- typed_function(
  function(x, y) x + y,
  params  = list(x = Numeric, y = Numeric),
  .return = Numeric
)

add(1, 2)     # Returns 3
add("a", 2)   # Error: Type error in 'x'
```


```{r, error = TRUE}
Num_not_inf <- create_type(
  name = "Num_not_inf", 
  checker = function(x) {
    is.numeric(x) && !is.infinite(x)
  }
)


divide <- function(a, b) {
  return(a / b)
}

divide_safe <- typed_function(
  fn = divide, 
  params = list(
    a = Num_not_inf, 
    b = Num_not_inf
  ), 
  .return = Num_not_inf
)


divide_safe(10, 2) # works normally
divide_safe(10, 0) # fails as 10/0 returns Inf
```


```{r, error = TRUE}
# Optional parameter
greet <- typed_function(
  function(name, title = NULL) {
    if (is.null(title)) paste("Hello,", name)
    else paste("Hello,", title, name)
  },
  params = list(name = String, title = Optional(String))
)

greet("Alice")                   # "Hello, Alice"
greet("Alice", title = "Dr.")    # "Hello, Dr. Alice"
greet("Alice", title = 42)       # Error: Type error in 'title'
```

```{r, error = TRUE}
# Union type in params
describe <- typed_function(
  function(id) paste("ID:", id),
  params  = list(id = String | Numeric),
  .return = String
)

describe("abc")   # "ID: abc"
describe(123)     # "ID: 123"
describe(TRUE)    # Error: Type error in 'id'
```

You can also define an object to expect as a return value: 


```{r, error = TRUE}
Person <- create_list_type(
  type_spec = list(
    name = String, 
    age = Numeric
  )
)

get_person_info_as_list <- function(name, age) {
  return(list(
    name = name, 
    age = age
  ))
}

get_person_info_as_message <- function(name, age) {
  return(
    paste("Hi my name is ", name, " I'm ", age, " years old")
  )
}

get_person_info_as_list_safe <- typed_function(
  fn = get_person_info_as_list, 
  params = list(name = String, age = Numeric), 
  .return = Person
)

get_person_info_as_list_safe(name = "Omar", age = 30) # works fine

get_person_info_as_message_safe <- typed_function(
  fn = get_person_info_as_message, 
  params = list(name = String, age = Numeric), 
  .return = Person
)

# Should fail as the function does not return a Person list anymore
get_person_info_as_message_safe(name = "Omar", age = 30) 

```


## 🌍 Function usage

```{r, error = TRUE}
# Catch bad payroll data early instead of getting silent NAs
calculate_mean_payroll <- function(salaries) {
  salaries %:% Numeric %<-% salaries
  mean(salaries)
}

calculate_mean_payroll(c(1800, 2300, 4000))   # Works fine

calculate_mean_payroll(c(1800, "2300", 4000)) # Error: type mismatch
```

## 📚 Learn More

Full documentation and worked examples are available at the [package website](https://feddelegrand7.github.io/sicher/).

